Post-replication citation patterns in psychology: Four case studies

### 2021-04-27 manuscript/appendix.Rmd
round 1 revisions added

```{r protocolAmendments}
tibble(
  "Issue" = c("Sampling frame", "Definition of pre-replication assessment period", "Exclusion of replication year from qualitative assessment", "Automated extraction of citation contexts and replication co-citation", "Counter-arguments", "Dual-coding for citation valence classification", "Examination of characteristics of articles providing counter-arguments.","Renaming of two citation patterns", "Data exclusions"),
  "Amendment" = c("The end date of the sampling frame was initially April 6th 2018 (see pre-registered protocol). Later, we decided to extend the sampling frame to April 26th 2019 to learn more about how the citation curves developed over time (see amended protocol). Finally, we decided to extend the sampling frame to December 31st 2019 to ensure we had all citations for the year 2019.", "Initially, the pre-replication assessment period was defined as 2015 (i.e., replication year - 1) for the Registered Replication Report cases and all pre-replication years for the other cases. In practice, when the replication year - 1 definition is used for all cases, only one study (Caruso) has a citing article outside of this period, and it is only a single article. Therefore, we have decided that it is more straightforward to use the same definition of the pre-replication assessment period (replication year - 1) for all cases.", "The pre-registered protocol neglected to mention that we intended to exclude the replication year from qualitative assessment. This change was documented in the amended protocol.", "Initially, we intended to use Matlab code to partly automate the extraction of ‘citation contexts’ (text surrounding in-text citations) and identify replication co-citation. However, we found that the size of the relevant citation context varied substantially between articles, so we decided to extract all citation contexts manually. We also decided to identify replication co-citation manually too. These changes were documented in the amended protocol.","In the amended protocol, we clarified that we would attempt to identify counter-arguments in all articles that cited the original study and the replication study.", "Initially, we planned to dual code small batches of articles depending on the degree of inter-rater reliability. However, after encountering a number of somewhat ambiguous cases in our initial coding, we decided to dual-code all articles in order to reduce subjectivity and increase consistency between coders. This is documented in the amended protocol.", "After completing the study we thought it would be informative to examine the article types and journals of articles that provided counter-arguments, and any overlap in authorship with the original study authors or prior collaborators of the first authors of the original studies. These analyses were not pre-registered.", "After completing the study we decided to revise the following terminology from the protocol in order to improve clarity: “principled defence” was changed to “explicit defence” and “hollow defence” was changed to “absent defence”.", "The criteria to exclude data from qualitative assessment (no access, non-English language, and no original study citation in the main text, were not pre-registered. These exclusions were necessary because they precluded the qualitative assessment.")
) %>%
  kable(booktabs=T,longtable=T,caption = "Protocol amendments.") %>%
  column_spec(1,width='4cm') %>%
  column_spec(2,width='11cm')
```

@@

>>>>

```{r protocolAmendments}
tibble(
  "Issue" = c("Sampling frame", "Definition of pre-replication assessment period", "Exclusion of replication year from qualitative assessment", "Automated extraction of citation contexts and replication co-citation", "Counter-arguments", "Dual-coding for citation valence classification", "Examination of characteristics of articles providing counter-arguments.","Renaming of two citation patterns", "Data exclusions", "Citations to replication studies and co-citation of original studies"),
  "Amendment" = c("The end date of the sampling frame was initially April 6th 2018 (see pre-registered protocol). Later, we decided to extend the sampling frame to April 26th 2019 to learn more about how the citation curves developed over time (see amended protocol). Finally, we decided to extend the sampling frame to December 31st 2019 to ensure we had all citations for the year 2019.", "Initially, the pre-replication assessment period was defined as 2015 (i.e., replication year - 1) for the Registered Replication Report cases and all pre-replication years for the other cases. In practice, when the replication year - 1 definition is used for all cases, only one study (Caruso) has a citing article outside of this period, and it is only a single article. Therefore, we have decided that it is more straightforward to use the same definition of the pre-replication assessment period (replication year - 1) for all cases.", "The pre-registered protocol neglected to mention that we intended to exclude the replication year from qualitative assessment. This change was documented in the amended protocol.", "Initially, we intended to use Matlab code to partly automate the extraction of ‘citation contexts’ (text surrounding in-text citations) and identify replication co-citation. However, we found that the size of the relevant citation context varied substantially between articles, so we decided to extract all citation contexts manually. We also decided to identify replication co-citation manually too. These changes were documented in the amended protocol.","In the amended protocol, we clarified that we would attempt to identify counter-arguments in all articles that cited the original study and the replication study.", "Initially, we planned to dual code small batches of articles depending on the degree of inter-rater reliability. However, we encountered a number of somewhat ambiguous cases in our initial coding which suggested there may be more subjectivity involved in the classification of citation valence than we had originally anticipated. For example, an article may contain several favourable statements and several unfavourable statements, in which case one needs to determine how to appropriately weight those statements in order to arrive at a classification that accurately represents the author’s viewpoint. To address this, we decided to minimize subjectivity by dual-coding all articles, with disagreements resolved through discussion (with a third coder if necessary). This is documented in the amended protocol.", "After completing the study we thought it would be informative to examine the article types and journals of articles that provided counter-arguments, and any overlap in authorship with the original study authors or prior collaborators of the first authors of the original studies. These analyses were not pre-registered.", "After completing the study we decided to revise the following terminology from the protocol in order to improve clarity: “principled defence” was changed to “explicit defence” and “hollow defence” was changed to “absent defence”.", "The criteria to exclude data from qualitative assessment (no access, non-English language, and no original study citation in the main text, were not pre-registered. These exclusions were necessary because they precluded the qualitative assessment.","Upon reviewer request, we obtain citation counts for replication studies and checked if they co-cited the relevant original study. These results are presented in Supplementary Information D.")
) %>%
  kable(booktabs=T,longtable=T,caption = "Protocol amendments.") %>%
  column_spec(1,width='4cm') %>%
  column_spec(2,width='11cm')
```

@@

# Citations to replication studies and co-citation of original studies {#appD}

Upon reviewer request, we examined citation counts for replication studies and checked whether citing articles also co-cited the relevant original study. To obtain the data, we downloaded bibliographic records from the Clarivate Analytics Web of Science Core Collection accessed via the University of Amsterdam on 16th April, 2021 for articles that cited each replication study (up to the study endpoint - December, 2019) and cross-checked them with our sample of articles that cited the original study. As shown, the replication studies also have a life of their own and they are often cited independently of the specific original study. Often this is easy to explain, e.g., Klein et al. replicated 13 original studies, not just the two that were of interest in our analysis. These studies are also likely to have accrued citations by virtue of being amongst the first highly prominent examples of preregistered multi-laboratory replications in psychology.

```{r}
# identify how many studies citing the replication also cite the original study

## extract citing articles for replication studies and original studies
d_kleinCites <- d_repCitations %>% filter(case == 'klein')
d_wagenmakersCites <- d_repCitations %>% filter(case == 'wagenmakers')
d_haggerCites <- d_repCitations %>% filter(case == 'hagger')
d_baumeisterCites <- d_citations %>% filter(case == 'baumeister')
d_sripadaCites <- d_citations %>% filter(case == 'sripada')
d_carusoCites <- d_citations %>% filter(case == 'caruso')
d_carterCites <- d_citations %>% filter(case == 'carter')
d_strackCites <- d_citations %>% filter(case == 'strack')

## compute citation counts for replication studies
hagCites <- nrow(d_haggerCites)
wagenCites <- nrow(d_wagenmakersCites)
kleinCites <- nrow(d_kleinCites)

## compute how many articles citing the replication study also cite the relevant original study
hagBauCocites <- sum(d_haggerCites$UT %in% d_baumeisterCites$UT)
hagSriCocites <- sum(d_haggerCites$UT %in% d_sripadaCites$UT)
wagenStrackCocites <- sum(d_wagenmakersCites$UT %in% d_strackCites$UT)
kleinCartCocites <- sum(d_kleinCites$UT %in% d_carterCites$UT)
kleinCaruCocites <- sum(d_kleinCites$UT %in% d_carusoCites$UT)

## build table
repCociteTable <- data.frame(
  "Replication" = c("Hagger et al. (2016)", "Klein et al. (2014)", "Wagenmakers et al. (2016)"),
  "Citation" = c(hagCites, kleinCites, wagenCites),
  "Cocitations" = c(paste0(hagBauCocites," (Baumeister et al. 1998) ",hagSriCocites," (Sripada et al. 2014)"), paste0(kleinCartCocites," (Carter et al. 2011) ",kleinCaruCocites," (Caruso et al. 2012)"), paste0(wagenStrackCocites," (Strack et al. 1988)"))
  )
```

```{r}
repCociteTable %>% kable(col.names = c("Replication study", "Citation count","Cocitations of original study"), caption = 'Citations counts for replication studies and co-citation counts to relevant original studies.', booktabs=T)
```

### 2021-04-27 manuscript/manuscript.Rmd
round 1 revisions added

---
title             : "Post-replication citation patterns in psychology: Four case studies"
shorttitle        : "Post-replication citation patterns"

@@

Table \@ref(tab:patterns) outlines several citation patterns that might follow a contradictory replication result, each reflecting different types of response by the research community. We have tentatively categorised these response patterns as being ‘progressive’ or ‘regressive’ depending on their expected impact on the accumulation of scientific knowledge[^footnote2]. The first set of patterns, belief correction/perpetuation, refer to what is often considered a primary functional role of (contradictory) replication studies - to change belief in the credibility of exaggerated or erroneous original findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018). In the absence of an explicit defence of an original study that convincingly explains a strongly contradictory replication result (see ‘explicit defence’ below), a progressive response might involve a decrease in favourable citations and an increase in unfavourable citations, reflecting updated beliefs about the credibility of the original finding. Conversely, a regressive response might involve maintenance of (or even increase in) favourable citations, and relatively few unfavourable citations, suggesting a perpetuation of belief in the credibility of the original finding despite the contradictory replication result. Prior research has documented how favourable citations to observational epidemiology studies can persist despite the claims of those studies being strongly contradicted in subsequent randomized trials (Tatsioni et al., 2007). Similarly, it has been reported that even when articles are retracted, they can continue to receive favourable citations (Budd et al., 1998). Thus, there is evidence that belief in the credibility of original findings can perpetuate even when subsequent events cast doubt on their credibility; however, we are unaware of similar evidence in the context of replication studies that were explicitly designed to test the reliability of prior findings.

@@

In the present study, we explored the post-replication citation patterns described above in the context of four case studies in the field of psychology where the findings of a replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)). In two of the cases, the replication studies were part of a single ‘Many Labs’ project (Klein et al., 2014) and addressed the ‘flag priming effect’ (T. J. Carter et al., 2011) and ‘money priming effect’ (Caruso et al., 2013) respectively. The other two cases involved ‘Registered Replication Reports’ (Simons et al., 2014), examining influential demonstrations of the ‘facial feedback effect’ (cf. Strack et al., 1988; Wagenmakers et al., 2016) and the ‘ego-depletion effect’ (cf. Baumeister et al., 1998; Hagger et al., 2016; Sripada et al., 2014) respectively. For methodological reasons (see Hagger et al., 2016), the ego-depletion replication was aimed at a classic study in the field (Baumeister et al., 1998), but actually employed a modified computer-based version of the original paradigm (Sripada et al., 2014). For this particular case study we examined citation patterns to both of these original studies. We chose these particular case studies because they involved prominent pre-registered multi-laboratory replication attempts with sample sizes between 23 - 211 times larger than the original studies, thus providing highly visible and highly credible evidence that strongly contradicted and outweighed earlier findings[^footnote3]. This facilitates additional interpretative clarity about the citation patterns one might expect to observe.

[^footnote3]: We deliberately focused on a select group of case studies rather than other potentially larger samples in order to aid interpretative clarity; for example, the extent to which the results of the large scale Reproducibility Project in Psychology (RPP; Open Science Collaboration, 2015) actually contradicted the original studies has been contested (Etz & Vandekerckhove, 2016). Additionally, original studies were not actually cited in the RPP research report, which may have diluted awareness about relevant replication studies.

@@

## Sample
We examined four case studies in which a prominent pre-registered and multi-laboratory replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)).

@@

Discussion
It has been proposed that replication studies can facilitate scientific self-correction by modifying scientists’ belief in the credibility of published findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018); however, the extent to which this occurs in practice is unclear. In this study, we investigated how the research community responded to four strongly contradictory replications in the field of psychology by examining post-replication citation patterns for original studies. We observed some ‘progressive’ response in the form of modest active correction patterns (a small decline in favourable citations and a small increase in unfavourable citations), but also even more prominent ‘regressive’ responses in the form of unchallenged belief perpetuation (sustained levels of favourable citations and few unfavourable citations, particularly in the Baumeister case) and considerable citation bias (neglecting to cite the replication study; in all cases aside from Sripada). When authors cited the original study favourably despite the replication result, only half of the articles provided any explicit counter arguments in defence of the original study (an explicit defence). Overall, these findings are consistent with prior observations that favourable citation patterns appear relatively unperturbed by subsequent publication of contradictory results in studies with lower risk of bias (Tatsioni et al., 2007) or even by full retraction (Budd et al., 1998), and that positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence once a theory gets entrenched despite overwhelming evidence against it (Bastiaansen et al., 2015; Greenberg, 2009).

It is reasonable to question whether the replication results in these case studies should (rationally) have instigated belief change in the research community and thus triggered a more sizable decline in favourable citations than we observed. It is important to note that the case studies we selected were deliberately chosen because the replication results were superior to the original results in terms of both credibility and evidential value. The replications were pre-registered, muli-laboratory studies with large sample sizes. By contrast, the original studies had much smaller sample sizes, and arguably had a much higher risk of bias, as they were not pre-registered, were performed by single teams, and arose in domains affected by publication bias and other questionable research practices (E. C. Carter et al., 2015; Coles et al., 2019; Vadillo, 2019; Vadillo et al., 2016). Thus, it seems reasonable to suggest that the compelling replication results should have reduced belief in the credibility of the original results.

The clear evidence of citation bias that our study documents may have two main contributory factors: (1) a lack of awareness about the replication results; and/or (2) a decision to ignore the replication results. The practical issue of awareness is not necessarily straightforward to address. Individual scientists can find it difficult to keep up-to-date with the voluminous literature that is relevant to their research. Recently, the reference manager Zotero has introduced a feature that alerts users when an article in their database has been retracted (Zotero, 2019). One could imagine a similar feature being introduced for replication studies, perhaps based on databases that explicitly identify replication studies such as Curate Science (https://curatescience.org/). However, it is much less straightforward to define and identify a relevant replication study (Neuliep & Crandall, 1993) and users would need to be alerted that this requires some scientific judgement rather than simple article meta-data. Another solution could be to encourage researchers to focus less on individual studies and more on up-to-date evidence summaries (i.e., reviews and meta-analyses) in which relevant evidence is systematically identified and collated. This would require that high-quality and contemporary evidence summaries are available; however, in psychology systematic reviews and meta-analyses can be of low quality and their results may still be inflated and non-reproducible (Kvarven et al., 2020; Maassen et al., 2020; Polanin et al., 2020). Moreover, empirical studies are often not included in any form of evidence synthesis (Hardwicke et al., 2020).

The second issue of authors ignoring highly relevant replications is undesirable and implies a biased appraisal of the evidence. Contradictory replication results do not necessarily nullify the credibility of an original study (Collins, 1985; Earp & Trafimow, 2015; Maxwell et al., 2015), but we would still expect highly relevant replication results to be cited and explicitly debated. In fact, we found that when authors continued to favourably cite original studies despite the replication results, around half did not provide explicit counter-arguments in defence of the original study. Although our study did not examine researchers’ individual beliefs, one recent study reported that when research psychologists were confronted with replication evidence, they often did update their (self-reported) beliefs, albeit modestly (McDiarmid et al., 2021). However, there are several reasons to be uncertain about whether the results from this artificial setting might generalize to real-world settings, including potential observer bias (i.e., participants behaving differently because they are under observation and/or responding to the perceived expectations of the research team) and the possibility that individuals may behave differently in settings where they have substantial personal investment and may be publicly scrutinized. Additionally, cognitive psychology studies have obtained some evidence of a ‘continued influence effect’ wherein an individual's beliefs and behaviour can continue to be influenced by false or misleading information despite subsequent efforts to reject it (Lewandowsky et al., 2012). It is plausible that various cognitive biases, such as confirmation bias (preferentially seeking out and processing evidence that supports pre-existing beliefs) or motivated reasoning (constructing and evaluating arguments based on what is desirable rather than what is rationally justifiable), may partly explain researchers' tendency to ignore or dismiss the replication evidence (Bishop, 2019; Kunda, 1990; Nickerson, 1998).

@@

We have only been able to gauge reactions to replications to the extent that they are reflected in citation patterns. Thus, for example, our method would not detect if there had been a correction effect amongst individuals who would not typically cite the original study, such as students, members of the public, or researchers working in other fields. A contested study may continue to be cited favourably by its proponents who remain working in the field. This will suffice to create belief perpetuation in the published literature, even though other scientists may simply no longer be interested in getting involved with such a strongly contested research topic. Relatedly, we did not examine citation patterns beyond 3-5 years post-replication. Some perspectives envision scientific self-correction unfolding over a much longer time scale (Lauden, 1981; Peterson & Panofsky, 2020) and suggest the process is characterised less by the impact of individual study results and more by the gradual accumulation of converging evidence, gradual revision of theoretical understanding, and/or informal sociological processes (e.g., researchers choosing alternative topics to study). Thus, whilst the current findings may contradict the expectations of a more direct and expedited view of the corrective impact of replication studies (Ioannidis, 2012; Vazire & Holcombe, 2020), they are not necessarily inconsistent with a slower and more indirect process of self-correction.

@@

Baumeister, R. E., Bratslavsky, E., Muraven, M., & Tice, D. M. (1998). Ego depletion: Is the active self a limited resource? Journal of Personality and Social Psychology, 74(5), 1252–1265. https://doi.org/10.1037/0022-3514.74.5.1252

@@

@@

@@

Hardwicke, T. E., Thibault, R. T., Kosie, J., Wallach, J. D., Kidwell, M. C., & Ioannidis, J. (2021). Estimating the prevalence of transparency and reproducibility-related research practices in psychology (2014-2017). Perspectives on Psychological Science. https://doi.org/10.31222/osf.io/9sz2y

@@

>>>>

---
title             : "Citation patterns following a strongly contradictory replication result: Four case studies from psychology"
shorttitle        : "Post-replication citation patterns"

@@

Table \@ref(tab:patterns) outlines several citation patterns that might follow a contradictory replication result, each reflecting different types of response by the research community. We have tentatively categorised these response patterns as being ‘progressive’ or ‘regressive’ depending on their expected impact on the accumulation of scientific knowledge[^footnote2]. The first set of patterns, belief correction/perpetuation, refer to what is often considered a primary functional role of (contradictory) replication studies - to change belief in the credibility of exaggerated or erroneous original findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018). In the absence of an explicit defence of an original study that convincingly explains a strongly contradictory replication result (see ‘explicit defence’ below), a progressive response might involve a decrease in favourable citations and an increase in unfavourable citations, reflecting updated beliefs about the credibility of the original finding. Conversely, a regressive response might involve maintenance of (or even increase in) favourable citations, and relatively few unfavourable citations, suggesting a perpetuation of belief in the credibility of the original finding despite the contradictory replication result. Prior research has documented how favourable citations to observational epidemiology studies can persist despite the claims of those studies being strongly contradicted in subsequent randomized trials (Tatsioni et al., 2007). Similarly, it has been reported that even when articles are retracted, they can continue to receive favourable citations (Budd et al., 1998; Fernandez & Vadillo, 2020). Thus, there is evidence that belief in the credibility of original findings can perpetuate even when subsequent events cast doubt on their credibility; however, we are unaware of similar evidence in the context of replication studies that were explicitly designed to test the reliability of prior findings.

@@

In the present study, we explored the post-replication citation patterns described above in the context of four case studies in the field of psychology where the findings of a replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)). In two of the cases, the replication studies were part of a single ‘Many Labs’ project (Klein et al., 2014) and addressed the ‘flag priming effect’ (T. J. Carter et al., 2011) and ‘money priming effect’ (Caruso et al., 2013) respectively. The other two cases involved ‘Registered Replication Reports’ (Simons et al., 2014), examining influential demonstrations of the ‘facial feedback effect’ (cf. Strack et al., 1988; Wagenmakers et al., 2016) and the ‘ego-depletion effect’ (cf. Baumeister et al., 1998; Hagger et al., 2016; Sripada et al., 2014) respectively. For methodological reasons (see Hagger et al., 2016), the ego-depletion replication was aimed at a classic study in the field (Baumeister et al., 1998), but actually employed a modified computer-based version of the original paradigm (Sripada et al., 2014). For this particular case study, we examined citation patterns to both of these original studies.

We adopted a case study approach in order to develop a ‘narrow and deep’ understanding of the topic, as opposed to ‘broad and shallow’ approach, which would have required a deliberate representative sampling strategy. We chose these particular case studies because they involved prominent pre-registered multi-laboratory replication attempts with sample sizes between 23 - 211 times larger than the original studies, thus providing highly visible and highly credible evidence that strongly contradicted and outweighed earlier findings[^footnote3]. This facilitates additional interpretative clarity about the citation patterns one might expect to observe.

[^footnote3]: We deliberately focused on a select group of case studies rather than other potentially larger samples in order to aid interpretative clarity; for example, the extent to which the results of the large-scale Reproducibility Project in Psychology (RPP; Open Science Collaboration, 2015) actually contradicted the original studies has been contested (Etz & Vandekerckhove, 2016). Additionally, original studies were not actually cited in the RPP research report, which may have diluted awareness about relevant replication studies.

@@

## Sample
We examined four case studies in which a prominent pre-registered and multi-laboratory replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)). As shown, all the original studies found modestly large to very large effects and all of them were relatively small studies (thus they would have been underpowered to detect small effects). Conversely, all the replication efforts comprised of very large sample sizes, and they would be very well powered to detect even small effects; however, they all obtained null results.

@@

Discussion
It has been proposed that replication studies can facilitate scientific self-correction by modifying scientists’ belief in the credibility of published findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018); however, the extent to which this occurs in practice is unclear. In this study, we investigated how the research community responded to four strongly contradictory replications in the field of psychology by examining post-replication citation patterns for original studies. We observed some ‘progressive’ response patterns in the form of modest active correction patterns (a small decline in favourable citations and a small increase in unfavourable citations), but also even more prominent ‘regressive’ response patternsß in the form of unchallenged belief perpetuation (sustained levels of favourable citations and few unfavourable citations, particularly in the Baumeister case) and considerable citation bias (neglecting to cite the replication study; in all cases aside from Sripada). When authors cited the original study favourably despite the replication result, only half of the articles provided any explicit counter arguments in defence of the original study (an explicit defence). Overall, these findings are consistent with prior observations that favourable citation patterns appear relatively unperturbed by subsequent publication of contradictory results in studies with lower risk of bias (Tatsioni et al., 2007) or even by full retraction (Budd et al., 1998; Fernandez & Vadillo, 2020), and that positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence once a theory gets entrenched, despite overwhelming evidence against it (Bastiaansen et al., 2015; Greenberg, 2009).

It is reasonable to question whether the replication results in these case studies should (rationally) have instigated belief change in the research community and thus triggered a more sizable decline in favourable citations than we observed. It is important to note that the case studies we selected were deliberately chosen because the replication results were superior to the original results in terms of both credibility and evidential value. The replications were pre-registered, muli-laboratory studies with large sample sizes. By contrast, the original studies had much smaller sample sizes, and arguably had a much higher risk of bias, as they were not pre-registered, were performed by single teams, and arose in domains affected by publication bias and other questionable research practices (E. C. Carter et al., 2015; Coles et al., 2019; Vadillo, 2019; Vadillo et al., 2016). Thus, it seems reasonable to suggest that the compelling replication results should have reduced belief in the credibility of the original results. However, one could also contend that if there were some flaw in the replication study that undermined its validity (Brandt et al., 2014; Vazire et al., 2021; Fabrigar et al., 2020), then this may provide justification to continue favourably citing the original study. Indeed, in these particular case studies, the validity of the replication results has been challenged by proponents of the original findings (Table 5; Baumeister & Vohs, 2016; Strack, 2016). As this debate remains far from settled (Coles et al., 2019; Vadillo, 2019), ideally any favourable citation of the original studies should at a minimum be accompanied by co-citation of the replication results and some discussion of the discrepant findings.

The clear evidence of citation bias that our study documents may have two main contributory factors: (1) a lack of awareness about the replication results; and/or (2) a decision to ignore the replication results. The practical issue of awareness is not necessarily straightforward to address. Individual scientists can find it difficult to keep up-to-date with the voluminous literature that is relevant to their research. Recently, the reference manager Zotero has introduced a feature that alerts users when an article in their database has been retracted (Zotero, 2019). One could imagine a similar feature being introduced for replication studies, perhaps based on databases that explicitly identify replication studies. However, it is much less straightforward to define and identify a relevant replication study (Neuliep & Crandall, 1993) and users would need to be alerted that this requires some scientific judgement rather than simple article meta-data. Another solution could be to encourage researchers to focus less on individual studies and more on up-to-date evidence summaries (i.e., reviews and meta-analyses) in which relevant evidence is systematically identified and collated. This would require that high-quality and contemporary evidence summaries are available; however, in psychology systematic reviews and meta-analyses can be of low quality and their results may still be inflated and non-reproducible (Kvarven et al., 2020; Maassen et al., 2020; Polanin et al., 2020). Moreover, empirical studies are often not included in any form of evidence synthesis (Hardwicke et al., 2020).

The second issue of authors ignoring highly relevant replications seems undesirable and implies a biased appraisal or presentation of the evidence. Contradictory replication results do not necessarily nullify the credibility of an original study (Collins, 1985; Earp & Trafimow, 2015; Maxwell et al., 2015), but we would still expect highly relevant replication results to be cited and explicitly debated. In fact, we found that when authors continued to favourably cite original studies despite the replication results, around half did not provide explicit counter-arguments in defence of the original study. Although our study did not examine researchers’ individual beliefs, one recent study reported that when research psychologists were confronted with replication evidence, they often did update their (self-reported) beliefs, albeit modestly (McDiarmid et al., 2021). However, there are several reasons to be uncertain about whether the results from this artificial setting might generalize to real-world settings, including potential participant reactivity (i.e., participants behaving differently because they are under observation and/or responding to the perceived expectations of the research team) and the possibility that individuals may behave differently in settings where they have substantial personal investment and may be publicly scrutinized. Additionally, cognitive psychology studies have obtained some evidence of a ‘continued influence effect’ wherein an individual's beliefs and behaviour can continue to be influenced by false or misleading information despite subsequent efforts to reject it (Lewandowsky et al., 2012). It is plausible that various cognitive biases, such as confirmation bias (preferentially seeking out and processing evidence that supports pre-existing beliefs) or motivated reasoning (constructing and evaluating arguments based on what is desirable rather than what is rationally justifiable), may partly explain researchers' tendency to ignore or dismiss the replication evidence (Bishop, 2019; Kunda, 1990; Nickerson, 1998).

@@

We have only been able to gauge reactions to replications to the extent that they are reflected in relatively short-term citation patterns. A potential explanation of the apparently cursory treatment of the replication studies could be that researchers only became aware of them after their own research projects had begun and or even had been completed. If one of the original studies had been a key motivator for one’s own study, then it may be difficult to accommodate the strongly contradictory replication results. It may even be tempting to ignore them, or give them only superficial treatment. Moreover, researchers who are convinced that the replication study has squarely refuted the original, may no longer be interested in doing research on a topic where they see no future potential. Additionally, examination of citation patterns would not detect if there had been a correction effect amongst individuals who would not typically cite the original study, such as students, members of the public, or researchers working in other fields. A contested study may continue to be cited favourably by its proponents who remain working in the field. This will suffice to create belief perpetuation in the published literature, even though other scientists may simply no longer be interested in getting involved with such a strongly contested research topic. Relatedly, we did not examine citation patterns beyond 3-5 years post-replication. Some perspectives envision scientific self-correction unfolding over a much longer time scale (Lauden, 1981; Peterson & Panofsky, 2020) and suggest the process is characterised less by the impact of individual study results and more by the gradual accumulation of converging evidence, gradual revision of theoretical understanding, and/or informal sociological processes (e.g., researchers choosing alternative topics to study). Thus, whilst the current findings may contradict the expectations of a more direct and expedited view of the corrective impact of replication studies (Ioannidis, 2012; Vazire & Holcombe, 2020), they are not necessarily inconsistent with a slower and more indirect process of self-correction. Future research could employ longitudinal designs or older historical case studies in order to evaluate citation patterns unfolding over a longer time scale.

@@

Baumeister, R. F., Bratslavsky, E., Muraven, M., & Tice, D. M. (1998). Ego depletion: Is the active self a limited resource? Journal of Personality and Social Psychology, 74(5), 1252–1265. https://doi.org/10.1037/0022-3514.74.5.1252

Baumeister, R. F., & Vohs, K. D. (2016). Misguided effort with elusive implications. Perspectives on Psychological Science, 11(4), 574-575. https://doi.org/10.1177%2F1745691616652878

@@

Brandt, M. J., IJzerman, H., Dijksterhuis, A., Farach, F. J., Geller, J., Giner-Sorolla, R., Grange, J. A., Perugini, M., Spies, J. R., van ’t Veer, A. (2014). The Replication Recipe: What makes for a convincing replication? Journal of Experimental Social Psychology, 50, 217–24. https://doi.org/10.1016/j.jesp.2013.10.005.

@@

Fabrigar, L. R., Wegener, D. T., Petty, R. E. (2020). A validity-based framework for understanding replication in psychology. Personality and Social Psychology Review, 24(4), 316-344. https://doi.org/10.1177/1088868320931366.

Fernandez, L. M., & Vadillo, M. A. (2020). Retracted papers die hard: Diederik Stapel and the enduring influence of flawed science. PsyArXiv. https://doi.org/10.31234/osf.io/cszpy

@@

Hardwicke, T. E., Thibault, R. T., Kosie, J., Wallach, J. D., Kidwell, M. C., & Ioannidis, J. (2021). Estimating the prevalence of transparency and reproducibility-related research practices in psychology (2014-2017). Perspectives on Psychological Science. https://doi.org/10.1177%2F1745691620979806

@@

Strack, F. (2016). Reflection on the smiling registered replication report. Perspectives on Psychological Science, 11(6), 929–930. https://doi.org/10.1177/1745691616674460
